---
title: "COVID_19"
author: "Tim Cerelli"
date: "2024-06-18"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Final Project 2 Covid 19 Data Analysis

Here in this project we will be using the John Hopkins data on COVID19 cases and deaths. We will be analyzing, and plotting our data. 

```{r urls, include=TRUE, message=FALSE, warning=FALSE}

library(tidyverse)
library(lubridate)

# Base URL for the COVID-19 data repository
url_in <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/"

# List of filenames
file_names <- c(
  'time_series_covid19_confirmed_US.csv',
  'time_series_covid19_confirmed_global.csv',
  'time_series_covid19_deaths_US.csv',
  'time_series_covid19_deaths_global.csv',
  'time_series_covid19_recovered_global.csv'
)

# Construct URLs by combining base_url with each filename
urls <- paste0(url_in, file_names)

# Read CSV files directly from URLs
us_cases <- read_csv(urls[1])
global_cases <- read_csv(urls[2])
us_deaths <- read_csv(urls[3])
global_deaths <- read_csv(urls[4])
global_recovered <- read_csv(urls[5])

```

## Tidy the Data 

Next we will want to tidy the data making it better for our data analysis. We will be putting items such as date, cases, and deaths we will put in their own columns

There is infomation we don't need such as lat and long we will remove. 

```{r tidydata, include=TRUE, message=FALSE, warning=FALSE}

#Global cases
global_cases <- global_cases %>%
  pivot_longer(cols = -c('Province/State', 'Country/Region', Lat, Long),
               names_to = 'date',
               values_to = 'cases') %>%
  select(-c(Lat, Long)) %>%
  rename(province_state = 'Province/State', country_region = 'Country/Region')

#Global deaths
global_deaths <- global_deaths %>%
  pivot_longer(cols = -c('Province/State', 'Country/Region', Lat, Long),
               names_to = 'date',
               values_to = 'deaths') %>%
  select(-c(Lat, Long)) %>%
  rename(province_state = 'Province/State', country_region = 'Country/Region')

#Combining global cases and global deaths
global <- global_cases %>%
  full_join(global_deaths) %>% 
  mutate(date = mdy(date)) %>%
  filter(cases > 0)

#US cases
us_cases <- us_cases %>%
  pivot_longer(cols = -c(UID, 'iso2', 'iso3', code3, 'Admin2', FIPS, 'Province_State', 'Country_Region', Lat, Long_, 'Combined_Key'),
               names_to = 'date',
               values_to = 'cases') %>%
  select(-c(UID, FIPS, code3, 'iso2', 'iso3', Lat, Long_, 'Combined_Key')) %>%
  rename(county = Admin2, province_state = Province_State, country_region = Country_Region)

#US deaths
us_deaths <- us_deaths %>%
  pivot_longer(cols = -c(UID, 'iso2', 'iso3', code3, FIPS, 'Admin2', 'Province_State', 'Country_Region', Lat, Long_, 'Combined_Key', Population),
               names_to = 'date',
               values_to = 'deaths') %>%
  select(-c(UID, 'iso2', 'iso3', code3, FIPS, Lat, Long_, 'Combined_Key')) %>%
  rename(county = Admin2, province_state = Province_State, country_region = Country_Region, population = Population)

#Combining US cases and US deaths
us <- us_cases %>%
  full_join(us_deaths) %>%
  mutate(date = mdy(date)) %>%
  filter(cases > 0)

#Making Us by state
us_by_state <- us %>%
  group_by(province_state, country_region, date) %>%
  summarize(cases = sum(cases), deaths = sum(deaths), 
            population = sum(population)) %>%
  mutate(deaths_per_mil = deaths * 1000000 / population, cases_per_mil = cases * 1000000 / population) %>%
  select(province_state, country_region, date, cases, cases_per_mil, deaths_per_mil, population, deaths) %>%
  ungroup()

#US totals
us_totals <- us_by_state %>%
  group_by(country_region, date) %>%
  summarize(cases = sum(cases), deaths = sum(deaths),
            population = sum(population)) %>%
  mutate(deaths_per_mil = deaths * 1000000 / population, 
         cases_per_mil = cases * 1000000 / population,
         new_cases = cases - lag(cases),
         new_deaths = deaths - lag(deaths)) %>%
  select(country_region, date, cases, deaths, new_cases, new_deaths, deaths_per_mil, cases_per_mil, population) %>%
  ungroup()

#Adding new cases and new deaths per million
us_totals <- us_totals %>%
  mutate(new_cases_per_mil = new_cases * 1000000 / population, 
         new_deaths_per_mil = new_deaths * 1000000 / population) %>%
  select(everything())


```


## Plotting our Data 

Here we will visualize and analyze our data. 

```{r plots, include=TRUE, message=FALSE, warning=FALSE }

#US cases and deaths per million
us_totals %>%
  ggplot() +
  geom_line(aes(x = date, y = deaths_per_mil, color = 'deaths')) + 
  geom_line(aes(x = date, y = cases_per_mil, color = 'cases')) +
  scale_y_log10() +
  labs(title = 'Cases per Million and Deaths per Million in United States', 
       x = 'Date', y = 'Deaths and Cases per Million')  +
  scale_x_date(date_labels = '%m-%Y', date_breaks = '2 month') +
  theme(axis.text.x = element_text(angle = 330))


```

```{r US_Cases_Mil, include=TRUE, message=FALSE, warning=FALSE}

#US new cases and new deaths per million
us_totals %>%
  ggplot() +
  geom_point(aes(x = date, y = new_cases_per_mil, color = 'Cases')) +
  geom_point(aes(x = date, y = new_deaths_per_mil, color = 'Deaths')) +
  scale_y_log10() + 
  labs(title = 'United States New Cases and New Deaths per Million', x = 'Date', y = 'New Cases and Deaths per Mil Scaled by Log 10') +
  scale_x_date(date_labels = '%m-%Y', date_breaks = '2 month') +
  theme(axis.text.x = element_text(angle = 330))

```



```{r USAVACC, include=TRUE, message=FALSE, warning=FALSE }

#us model of deaths and cases per mil to predict death per mil
us_deaths_model <- lm(deaths_per_mil ~ cases_per_mil, data = us_totals)
summary(us_deaths_model)

us_totals <- us_totals %>%
  mutate(pred_deaths_per_mil = predict(us_deaths_model))
us_totals %>%
  ggplot() +
  geom_point(aes(x = cases_per_mil, y = deaths_per_mil, color = 'Actual Deaths per Million')) + 
  geom_point(aes(x = cases_per_mil, y = pred_deaths_per_mil, color = 'Predicted Deaths per Million')) +
  labs(title = 'American Deaths per Million as a Function of Cases per Million', x = 'Cases per Million', y = 'Deaths per Million')




```



