---
title: "NYC Shooting Analysis"
autor: Tim Cerelli
date: "2024-06-17"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Introduction

In this data analysis report we will analyze the historical shooting incidents in New York city. 
We will be cleaning and analyzing the data to create a forecast model of future results. 

My objective in this analysis is to see one the most dangerous neighborhoods based on shootings, and to analyze if these neighborhoods are getting safer. 
Also I would like to analyze the similarities between perpetrator sex and the sex of the victim 

#### Part 1 Overview of the Data
The data set contains historical shooting incidents in New York City. The key variables include:

OCCUR_DATE: The date of the shooting incident.

BORO: The borough where the incident occurred.

VIC_AGE_GROUP: The age group of the victim.

VIC_SEX: The sex of the victim.

VIC_RACE: The race of the victim.

PERP_AGE_GROUP: The age group of the perpetrator.

PERP_SEX: The sex of the perpetrator.


Import Libraries and Data
```{r packages, include=TRUE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Load necessary packages
library(dplyr)
library(tidyr)
library(ggplot2)
library(lubridate)


# URL for the dataset
url_in <- "https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD"

# Read data from CSV file
data <- read.csv(url_in)
```

##Part 2 Tidy Data

```{r clean, include=TRUE, message=FALSE, warning=FALSE}
# Clean and prepare the data
tidy_data <- data %>%
  select(-c(INCIDENT_KEY, JURISDICTION_CODE, LOC_CLASSFCTN_DESC, Latitude, Longitude, 
            Lon_Lat, X_COORD_CD, Y_COORD_CD, PRECINCT, LOCATION_DESC, LOC_OF_OCCUR_DESC, 
            STATISTICAL_MURDER_FLAG, OCCUR_TIME)) %>%
  filter(!is.na(PERP_SEX) & PERP_SEX %in% c("M", "F")) %>%
  mutate(OCCUR_DATE = as.Date(OCCUR_DATE, format = "%m/%d/%Y"),
         WEEK = lubridate::floor_date(OCCUR_DATE, "week"))

# Load necessary libraries
library(dplyr)

# Assuming tidy_data is your dataframe containing the shooting incident data

# Step 1: Identify values with fewer than 10 occurrences in VIC_AGE_GROUP
vic_age_counts <- tidy_data %>%
  count(VIC_AGE_GROUP) %>%
  arrange(n)

less_than_10_vic <- vic_age_counts %>%
  filter(n < 10)

# Step 2: Clean and transform VIC_AGE_GROUP to remove age groups that don't make sense
tidy_data_clean <- tidy_data %>%
  mutate(
    VIC_AGE_GROUP_CLEAN = case_when(
      VIC_AGE_GROUP %in% less_than_10_vic$VIC_AGE_GROUP ~ "Other",  # Group less frequent values
      TRUE ~ as.character(VIC_AGE_GROUP)                           # Keep other values as they are
    )
  )

# Step 3: Identify values with fewer than 10 occurrences in PERP_AGE_GROUP
perp_age_counts <- tidy_data %>%
  count(PERP_AGE_GROUP) %>%
  arrange(n)

less_than_10_perp <- perp_age_counts %>%
  filter(n < 10)

# Step 4: Clean and transform PERP_AGE_GROUP to remove age groups that don't make sense
tidy_data_clean <- tidy_data_clean %>%
  mutate(
    PERP_AGE_GROUP_CLEAN = case_when(
      PERP_AGE_GROUP %in% less_than_10_perp$PERP_AGE_GROUP ~ "Other",  # Group less frequent values
      PERP_AGE_GROUP %in% c("1020", "1022", "1028","other") ~ "10-29",       # Correct specific values
      PERP_AGE_GROUP == "224" ~ "18-24",                             # Correct another value
      TRUE ~ as.character(PERP_AGE_GROUP)                            # Keep other values as they are
    )
  )

```


### Part 2 Tidy, transform, and summarize statistics 
Let's start by looking at some summary statistics for the key variables.



```{r summary, include=TRUE, message=FALSE, warning=FALSE}

head(tidy_data)

# Convert relevant columns to factors
tidy_data$BORO <- as.factor(tidy_data$BORO)
tidy_data$VIC_AGE_GROUP <- as.factor(tidy_data$VIC_AGE_GROUP)
tidy_data$VIC_SEX <- as.factor(tidy_data$VIC_SEX)
tidy_data$VIC_RACE <- as.factor(tidy_data$VIC_RACE)

# Print the summary of tidy_data
summary(tidy_data)

```

```{r plottingdata, include=TRUE, message=FALSE, warning=FALSE}
# Aggregate data by year
yearly_data <- tidy_data %>%
  group_by(BORO, YEAR = lubridate::year(OCCUR_DATE)) %>%
  summarise(total_shootings = n()) %>%
  ungroup()


# Aggregate data by  month
monthly_data <- tidy_data %>%
  group_by(BORO, MONTH = floor_date(OCCUR_DATE, "month")) %>%
  summarise(total_shootings = n()) %>%
  ungroup()


# Aggregate data by borough and week
  weekly_data <- tidy_data %>%
    group_by(BORO, WEEK) %>%
    summarise(total_shootings = n()) %>%
    ungroup()
  
 # Pivot the table to wide format by perpetrator sex
  cases_by_date_sex <- tidy_data %>%
    filter(PERP_SEX %in% c("M", "F")) %>%
    group_by(OCCUR_DATE, PERP_SEX) %>%
    summarise(total_cases = n()) %>%
    pivot_wider(names_from = PERP_SEX, values_from = total_cases, values_fill = 0) %>%
    rename(M_PERP_SEX = M, F_PERP_SEX = F)
  
  

# Plotting the data as a line graph by year
plot1 <- ggplot(data = yearly_data, aes(x = YEAR, y = total_shootings, color = BORO)) +
  geom_line() +
  labs(title = "Yearly Number of Shootings by Borough",
       x = "Year",
       y = "Total Shootings",
       color = "Borough") +
  theme_minimal()

# Assuming monthly_data is already created as per the previous instructions

plotmonthly <- ggplot(data = monthly_data, aes(x = MONTH, y = total_shootings, color = BORO)) +
  geom_line() +
  labs(title = "Monthly Number of Shootings by Borough",
       x = "Time",
       y = "Total Shootings",
       color = "Borough") +
  theme_minimal()


# Plotting the data
  plot2 <- ggplot(data = weekly_data, aes(x = WEEK, y = total_shootings, color = BORO)) +
    geom_line() +
    labs(title = "Weekly Number of Shootings by Borough",
         x = "Week",
         y = "Total Shootings",
         color = "Borough") +
    theme_minimal()

# Bar plot of perpetrator age groups
plot_perp_age <- ggplot(tidy_data_clean, aes(x = PERP_AGE_GROUP_CLEAN)) +
  geom_bar(fill = "#4682B4") +
  labs(title = "Distribution of Perpetrator Age Groups",
       x = "Perpetrator Age Group",
       y = "Count") +
  theme_minimal()

# Bar plot of cleaned victim age groups
plot_vic_age <- ggplot(tidy_data_clean, aes(x = VIC_AGE_GROUP_CLEAN)) +
  geom_bar(fill = "#4682B4") +
  labs(title = "Distribution of Victim Age Groups",
       x = "Victim Age Group",
       y = "Count") +
  theme_minimal()
  
# Plotting the data
plot3 <- ggplot(data = cases_by_date_sex, aes(x = OCCUR_DATE)) +
  geom_line(aes(y = M_PERP_SEX, color = "Male Perpetrator")) +
  geom_line(aes(y = F_PERP_SEX, color = "Female Perpetrator")) +
  labs(title = "Number of Shootings by Perpetrator Sex Over Time",
       x = "Date",
       y = "Total Shootings",
       color = "Perpetrator Sex") +
  theme_minimal()


# Prepare data for logistic regression
logistic_data <- tidy_data %>%
  select(PERP_SEX, PERP_AGE_GROUP, BORO, VIC_AGE_GROUP, VIC_SEX) %>%
  mutate(PERP_SEX = ifelse(PERP_SEX == "M", 1, 0)) %>%
  na.omit()

# Fit logistic regression model
modelsex <- glm(PERP_SEX ~ PERP_AGE_GROUP + BORO + VIC_AGE_GROUP + VIC_SEX, 
             data = logistic_data, 
             family = binomial)

# Prepare data for regression
regression_data <- tidy_data %>%
  group_by(BORO, PERP_AGE_GROUP) %>%
  summarise(total_shootings = n()) %>%
  na.omit()

# Fit multiple linear regression model
modelshootings <- lm(total_shootings ~ BORO + PERP_AGE_GROUP, data = regression_data)

```

## Part 3 Plot Summary Analysis 

From our plots we can determine that perpetrators are more likely to be younger and male. It is also noticeable that there are certain neighborhoods that have a higher likelihood of shootings to be present. From our graphs you will see both the Bronx and Brooklyn have the highest percentage of shootings.  


```{r plots, echo=FALSE, message=FALSE, warning=FALSE}
#print(plot1)
#print(plot2)
print(plot3)
#print(plot_perp_age)
#print(plot_vic_age)
print(plotmonthly)
#summary(modelsex)
#summary(modelshootings)
```

## Part 3 Plot Summary Analysis 

From the pie graph you will see both the Bronx and Brooklyn have the highest percentage of shootings.  

```{r piechart, include=TRUE, echo=FALSE, message=FALSE, warning=FALSE}

# Load necessary packages

# Aggregate data by borough
borough_data <- tidy_data %>%  group_by(BORO) %>%  summarise(total_shootings = n()) %>%  ungroup() %>%  mutate(percentage = total_shootings / sum(total_shootings) * 100)

# Create a pie chart
ggplot(borough_data, aes(x = "", y = percentage, fill = BORO)) +  geom_bar(stat = "identity", width = 1) +  coord_polar(theta = "y") +  labs(title = "Percentage of Shootings by Borough",       x = "",       y = "",       fill = "Borough") +  theme_minimal() +  theme(axis.text.x = element_blank(),        axis.ticks = element_blank(),        panel.grid = element_blank()) +  geom_text(aes(label = paste0(round(percentage, 1), "%")),             position = position_stack(vjust = 0.5))



```

```{r forecast, echo=FALSE, message=FALSE, warning=FALSE}
# Install and load the forecast package
library(forecast)

# Aggregate data by year and borough
yearly_data <- ungroup(summarise(group_by(tidy_data, BORO, YEAR = lubridate::year(OCCUR_DATE)), total_shootings = n()))

# Get the list of unique boroughs
boroughs <- unique(yearly_data$BORO)

# Create an empty data frame to store forecasted values
forecasted_data <- data.frame()

# Loop through each borough and create a forecast
for (borough in boroughs) {
  # Filter data for the specific borough
  borough_data <- arrange(filter(yearly_data, BORO == borough), YEAR)  
  # Create a time series object
  ts_data <- ts(borough_data$total_shootings, start = min(borough_data$YEAR), frequency = 1)  
  # Fit an ARIMA model
  fit <- auto.arima(ts_data)  
  # Forecast the next 5 years
  forecasted_values <- forecast(fit, h = 5)  
  # Convert forecasted values to a data frame
  forecasted_df <- data.frame(    YEAR = as.numeric(time(forecasted_values$mean)),    total_shootings = as.numeric(forecasted_values$mean),    BORO = borough  )  
  # Combine with the original data
  combined_data <- rbind(borough_data, forecasted_df)  
  # Add to the forecasted data frame
  forecasted_data <- rbind(forecasted_data, combined_data)}

#print(forecasted_data)

# Plot the forecasted values for all boroughs
plot_forecast <- ggplot(forecasted_data, aes(x = YEAR, y = total_shootings, color = BORO)) +  geom_line() +  labs(title = "Forecasted Number of Shootings by Borough",       x = "Year",       y = "Total Shootings",       color = "Borough") + theme_minimal()
```

## Part 3 Data Model Analysis 

In this graph we are presenting the probability of a certain age group perpetrator within a given borough. The null hypothesis is that a borough does not have an impact on the age group of a shooter and I would like to see if a borough can affect this. 


I will be using a chi-squared test with a 0.05 significance level to test this. 



```{r boroughs, echo=FALSE, message=FALSE, warning=FALSE}
# Aggregate data to get proportions by borough and age group
agg_data <- tidy_data_clean %>%
  group_by(BORO, PERP_AGE_GROUP_CLEAN) %>%
  summarise(count = n()) %>%
  group_by(BORO) %>%
  mutate(prob = count / sum(count)) %>%
  ungroup()


# Plotting
plot_boroughs <- ggplot(agg_data, aes(x = PERP_AGE_GROUP_CLEAN, y = prob, fill = BORO)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Likelihood of Perpetrator Age Group by Borough",
       x = "Perpetrator Age Group",
       y = "Probability",
       caption = "Data source: NYPD Shooting Incidents") +
  theme_minimal() +
  theme(plot.title = element_text(size = 14, face = "bold"),
        axis.title = element_text(size = 12),
        axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "bottom") +
  scale_fill_brewer(palette = "Set3")

print(plot_boroughs)

# Example chi-squared test result (replace with your actual result)
chi_squared_test_result <- chisq.test(table(agg_data$BORO, agg_data$PERP_AGE_GROUP_CLEAN))

# Significance level
alpha <- 0.05

# Extract p-value from the test result
p_value <- chi_squared_test_result$p.value

print(p_value)

# Interpretation based on p-value and alpha
if (p_value < alpha) {
  cat("Reject the null hypothesis: There is a significant association between borough and age group.\n")
} else {
  cat("Fail to reject the null hypothesis: There is no significant association between borough and age group.\n")
}


# Print the chi-squared statistic and degrees of freedom
cat("\nChi-squared:", chi_squared_test_result$statistic, "\n")
cat("Degrees of freedom:", chi_squared_test_result$parameter, "\n")
```

## Part 4 Bias Identification

To conclude my project I found that there was no evidence to show a relation between perpetrator age being centrally located in a specific neighborhood. I thought the neighborhood might have an impact on the way age groups act. This was not to be the case as it was seen that it was just more likely a younger age group were to be the perpetrator no matter the neighborhood. I took out my bias by using statistics and sticking to the numbers to show what truly happened. 